Fully Extrinsic Type Inference (DRAFT 2025-06-20)

### 2025-12-23 doc/extrinsic_type_inference_paper/paper.tex
update static subtyping proof automation
update static subtyping solver for lfp elim
add proofs for union intro, inter elim, and universal elim
add soundness statement and update naming convetion
update lfp case to use full dynamic recursive call to ensure expressiveness
update annotation case for typing soundness
update stream case of loop typing soundness
restructure loop typing soudness with substance lemma
update notes on completeness nomenclature
add note on semantics-directed typing

      \delta \satisfies e \hastype \J{LFP[}\alpha\J{]}\tau 
      &\iff& 
      \alpha \wr^+ \tau
      \up
      \exists n \qua \VDash e \hastype (\tau[\alpha \slash \J{BOT}]^n)[\delta]

@@

@@

    \inferrule {
      \textbf{ftv}(\tau_a) \subseteq \emptyset
      \\
      \Theta, \Delta, \Gamma \entails e \hastype \Pi
      \\
      \textbf{pack}^+(\textbf{ftv}(\Delta) \entails \Pi) = \tau_e
      \\
      \exists \Theta', \Delta' \qua \Theta, \Delta \entails \tau_e \subtypes \tau_a \given \Theta', \Delta'
    } {
      \Theta, \Delta, \Gamma \entails e \J{ as } \tau_a \hastype \tau_a \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

    \inferrule {
      \alpha \neq \alpha_l
      \\\\
      \textbf{invert}(\alpha \entails \Delta) = \Delta' 
      \\
      \textbf{pack}^-(\alpha\ \alpha_l\ \Theta^\dagger \entails \Theta, \Delta', \alpha_l \J{*} \tau_r) = \tau'
      \\\\
      \textbf{factor}(\alpha \entails \tau' \cdot \J{left}) = \tau_l
      \\
      \textbf{factor}(\alpha \entails \tau' \cdot \J{right}) = \tau_r'
      \\\\
      \alpha_l \wr^+ \tau_r'
      \\
      % founded upper bound
      \alpha \downarrow \tau_l \subtypes \tau_l'
      \\
      \tau_r'[\alpha_l\slash \J{LFP[} \alpha \J{]} \tau_l'] = \tau_r''
    } {
      \Theta^\dagger \entails \alpha \circlearrowleft (\Theta, \Delta, \alpha_l \J{->}\tau_r) \subtypes \alpha_l \J{->} \J{LFP[} \alpha \J{]} \tau_r''
    }
  \end{mathpar}
\end{definition}

@@

\begin{definition} 
  \label{def:proof_subtyping_lfp_elimination}
  \emph{proof subtyping (least fixed point elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \J{LFP[}\alpha\J{]}\tau_l \subtypes \tau_r \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails 
      \textbf{factor}(\alpha \entails \tau_l \cdot l) 
      \subtypes 
      \tau_r \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \J{LFP[}\alpha\J{]}\tau_l \subtypes \J{<} l \J{>} \tau_r 
      \given \Theta', \Delta' 
    }

@@

    \inferrule {
      \Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'
      \\\\
      \Theta', \Delta' \entails \Delta_r \given \Theta'', \Delta'' 
    } {
      \Theta, \Delta \entails 
      \tau_l
      \subtypes 
      \J{EXI[}\Theta_r\J{]}\Delta_r \J{:} \tau_r \given \Theta'', \Delta'' 
    }
  \end{mathpar}
\end{definition}
\noindent
For union introduction \ms{
\tau \subtypes \tau_{l}\J{|}\tau_{r}
}, 
it can choose from two possible options:
solving the introduction of of left part, 
or solving the introduction of the right part.
For existential introduction \ms{ 
\tau_l \subtypes \J{EXI[}\Theta_r \J{]} \Delta_r \J{:}\tau_r
},
it constructs witnesses by solving the introduction
of the existential's constituent type \ms{\entails \tau_l \subtypes \tau_r \given}, 
from which the witnesses can be inferred from the generated subtyping constraints.
It then attempts to solve the qualifiers by leveraging the witnesses.

@@

    \inferrule {
      \Theta, \Delta \entails \textbf{merge}(\tau_{l}\J{\&}\tau_{r}) \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails \tau_{l}\J{\&}\tau_{r}  \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
    }

@@

\begin{theorem}
  \label{thm:proof_subtyping_substance}
  \emph{proof subtyping substance}
  \small
  \nopad
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'
    } {
      \forall \delta \qua \textbf{dom}(\delta) \subseteq \Theta \implies
      \exists \delta' \qua \delta \oplus \delta' \satisfies \Delta
      \implies
      \exists \delta' \qua \delta \oplus \delta' \satisfies \Delta'
    }
  \end{mathpar}
\end{theorem}

@@

\begin{theorem} 
  \label{thm:proof_subtyping_soundness}
  \emph{proof subtyping soundness} 
  \small
  \nopad
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'
    } {
      \exists \delta' \qua 
      \textbf{dom}(\delta') \subseteq \Theta' \backslash \Theta
      \up
      (
      \forall \delta \qua 
      \delta \oplus \delta' \satisfies \Delta' \implies
      \delta \oplus \delta' \satisfies \tau_l \subtypes \tau_r
      )
    }
  \end{mathpar}
\end{theorem}

>>>>

      \delta \satisfies e \hastype \J{LFP[}\alpha\J{]}\tau 
      &\iff& 
      \alpha \wr \tau
      \up
      \exists \tau' \qua
        \delta \satisfies \tau' \subtypes \tau \cdot \alpha \slash \tau'
        \up
        \alpha \slash \tau'\ \delta \satisfies e \hastype \tau

@@

\TODO{
RATIONALE -- SEMANTICS-DIRECTED TYPING:
note that we use semantics-directed typing (i.e. dynamic) rather than syntax-directed typing (i.e. static), in order
to avoid overly restricting the types of constructions like functions or records.
That is, in a syntax-directed semantics, we strive to have functions 
introduce the strongest type possible, which requires combining numerous type forms,
and parsing the structures of expressions.
To ensure the semantics of type forms is clear and intuitive, 
we use sementics-directed typing, which allows
each type form to be specified separately.
In essence, the semantics-directed typing is a denotation
from type syntax to predicate logic.
By contrast, simple type systems are so simple that the syntax-directed typing definitions
are intuitive already, and so there is no need for a separate semantics-directed definition.
}

@@

    \inferrule {
      \textbf{ftv}(\tau_a) \subseteq \emptyset
      \\
      \Theta, \Delta, \Gamma \entails e \hastype \tau_e \given \Theta', \Delta'
      \\
      \Theta', \Delta' \entails \tau_e \subtypes \tau_a \given \Theta'', \Delta''
    } {
      \Theta, \Delta, \Gamma \entails e \J{ as } \tau_a \hastype \tau_a \given \Theta'', \Delta''
    }
  \end{mathpar}
\end{definition}

@@

    \inferrule {
      \alpha \neq \alpha_l
      \\\\
      \textbf{invert}(\alpha \entails \Delta) = \Delta' 
      \\
      \textbf{pack}^-(\alpha\ \alpha_l\ \Theta^\dagger \entails \Theta, \Delta', \alpha_l \J{*} \tau_r) = \tau'
      \\\\
      \textbf{factor}(\alpha \entails \tau' \cdot \J{left}) = \tau_l
      \\
      \textbf{factor}(\alpha \entails \tau' \cdot \J{right}) = \tau_r'
      \\\\
      \alpha_l \wr^+ \J{LFP[} \alpha \J{]}\tau_r'
      \\
      % founded upper bound
      \alpha \downarrow \tau_l \subtypes \tau_l'
      \\
      (\J{LFP[} \alpha \J{]}\tau_r')[\alpha_l\slash \J{LFP[} \alpha \J{]} \tau_l'] = \tau_r''
    } {
      \Theta^\dagger \entails \alpha \circlearrowleft (\Theta, \Delta, \alpha_l \J{->}\tau_r) \subtypes \alpha_l \J{->} \tau_r''
    }
  \end{mathpar}
\end{definition}

@@

\begin{definition} 
  \label{def:proof_subtyping_lfp_elimination}
  \emph{proof subtyping (least fixed point elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \J{LFP[}\alpha\J{]}\tau_l \subtypes \tau_r \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \textbf{factor}(\alpha \entails \tau_l \cdot l) = \tau
      \\
      \Theta, \Delta \entails 
      \J{LFP[}\alpha\J{]} \tau
      \subtypes 
      \tau_r \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \J{LFP[}\alpha\J{]}\tau_l \subtypes \J{<} l \J{>} \tau_r 
      \given \Theta', \Delta' 
    }

@@

    \inferrule {
      \TODO{maybe learned subtypings on bound variabes can be removed}
      \\\\
      \Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'
      \\\\
      \Theta', \Delta' \entails \Delta_r \given \Theta'', \Delta'' 
    } {
      \Theta, \Delta \entails 
      \tau_l
      \subtypes 
      \J{EXI[}\Theta_r\J{]}\Delta_r \J{:} \tau_r \given \Theta'', \Delta'' 
    }
  \end{mathpar}
\end{definition}
\noindent
For union introduction \ms{
\tau \subtypes \tau_{l}\J{|}\tau_{r}
}, 
it can choose from two possible options:
solving the introduction of of left part, 
or solving the introduction of the right part.
For existential introduction \ms{ 
\tau_l \subtypes \J{EXI[}\Theta_r \J{]} \Delta_r \J{:}\tau_r
},
it constructs witnesses by solving the introduction
of the existential's constituent type \ms{\entails \tau_l \subtypes \tau_r \given}, 
from which the witnesses can be inferred from the generated subtyping constraints.
It then attempts to solve the qualifiers by leveraging the witnesses.

@@

    %% TODO: remove merge rule. Should be subsumed by antecedent union
    % \inferrule {
    %   \Theta, \Delta \entails \textbf{merge}(\tau_{l}\J{\&}\tau_{r}) \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
    % } {
    %   \Theta, \Delta \entails \tau_{l}\J{\&}\tau_{r}  \subtypes \tau_a \J{->} \tau_b \given \Theta', \Delta'
    % }

@@

\TODO{consider replacing "substance" with some variation on "conditional completemness"}
\begin{theorem}
  \label{thm:proof_subtyping_substance}
  \emph{proof subtyping substance}
  \small
  \nopad
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'
    } {
      \delta \satisfies \Delta
      \implies
      \exists \delta' \qua \delta \oplus \delta' \satisfies \Delta'
    }
  \end{mathpar}
\end{theorem}

@@

\begin{theorem} 
  \label{thm:proof_subtyping_soundness}
  \emph{proof subtyping soundness} 
  \small
  \nopad
  \begin{mathpar}
    \inferrule {
      \TODO{switch ordering of assignment maps}
      \\\\
      \TODO{look up from left to right}
      \\\\
      \Theta, \Delta \entails \tau_l \subtypes \tau_r \given \Theta', \Delta'
    } {
      \exists \delta' \qua 
      \textbf{dom}(\delta') \subseteq \Theta' \backslash \Theta
      \up
      (
      \forall \delta \qua 
      \delta \oplus \delta' \satisfies \Delta' \implies
      \delta \oplus \delta' \satisfies \tau_l \subtypes \tau_r
      )
    }
  \end{mathpar}
\end{theorem}
